\texttt{dispRity} demo for palaeobiologists

### 2015-12-01 doc/dispRity-manual.Rnw

<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Graphical options
par(bty = "n")

@@

@@

\section{Developments}
\label{developments}
As stated at the start of the demo, this version \Sexpr{packageVersion("dispRity")} is still in development and many parts are missing.
Here are the new functionalities that will be implemented before the proper release version (v.1).

\subsection{More user defined stuff}
I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via \texttt{make.boot}) or the evolutionary models (via \texttt{make.model}).
Both functions will provide similar testing as the \texttt{make.metric} function.

@@

>>>>

<<plot4, fig.width=8, fig.height=8, out.width='.5\\linewidth'>>=
## Graphical options
par(bty = "n")

@@

Finally, the \dispRity package allows to apply some tests to the data in order to properly test hypothesis.
The function \texttt{test.dispRity} works in a similar way as the \dispRity function: it intakes a \dispRity object (containing disparity measurements), a \texttt{test} and a \texttt{comparison} argument.
The \texttt{test} argument can be any statistical or non-statistical test to apply to the disparity object.
It can be common test function (e.g. \texttt{t.test::stats}) or some functions implemented in \dispRity (see \texttt{dispRity.test}) or even user defined functions.
The \texttt{comparison} argument must indicate the way the test should be applied to the data:
\begin{itemize}
\item \texttt{"pairwise"} (default): to compare each series pairwise
\item \texttt{"referential"}: to compare each series to the first one
\item \texttt{"sequential"}: to compare each series to the following one
\item \texttt{"all"}: to compare all the series together (like in analysis of variance)
\end{itemize}
It is also possible to input a list of pairs of \texttt{numeric} values or \texttt{characters} matching the series names to create personalised test.
This function also allows to correct for type I error inflation when using multiple comparisons via the \texttt{correction} argument.
This argument can be empty (no correction applied) or can contain one of the corrections from the \texttt{p.adjust::stats} function (see \texttt{?p.adjust}).

Note that the \texttt{test.dispRity} algorithm deals with some classic tests outputs (\texttt{h.test}, \texttt{lm} and \texttt{numeric} vector) and summarises the test output.
It is however possible to get the full detailed output by using the options \texttt{details = TRUE}.

<<test.dispRity>>=
## Performing a pairwise t-test to the test a difference in disparity between
## crown and stem mammals (note that "pairwise" is default)
test.dispRity(disp_crown_stemBS, test = t.test)

## The same test but with the detailed output
test.dispRity(disp_crown_stemBS, test = t.test, details = TRUE)

## A wilcoxon rank test applied to sliced disparity the data sequentially with
## the Bonferroni correction.
test.dispRity(disp_time_slicesBS, test = wilcox.test,
    comparisons = "sequential", correction = "bonferroni")

## An analysis of variance (aov) applied to the same data
test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")

## Note that in this case, the output is the regular aov output:
slice_aov <- test.dispRity(disp_time_slicesBS, test = aov, comparisons = "all")
class(slice_aov) ; summary(slice_aov)

## Measuring the overlap between distributions in the time bins (using the
## implemented Bhattacharyya Coefficient function - see ?bhatt.coeff) with a
## personalised series comparisons and some additional arguments to pass to the
## test function (bw = bw.nrd)
test.dispRity(disp_time_binsBS, test = bhatt.coeff, bw = bw.nrd,
    comparisons = list(c(1,2), c(1,3), c(2,1), c(3,1)))

@@

\section{Developments}
\label{developments}
As stated at the start of the demo, this version \Sexpr{packageVersion("dispRity")} is still in development and many parts are missing.
Here are the new functionalities that will be implemented in further versions.

\subsection{More user defined functions}
I intend also develop functions to help users to develop their own algorithms for the bootstrap method (via \texttt{make.boot}) or the evolutionary models (via \texttt{make.model}).
Both functions will provide similar testing as the \texttt{make.metric} function.

@@

\subsection{Suggestions?}
If you have any extra suggestions or comments on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{guillert@tcd.ie}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/dispRity}{GitHub page (github.com/TGuillerme/dispRity)}.

### 2015-12-01 doc/dispRity-palaeo-demo.Rnw

This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in this dataset since the last 100 million years.

@@

<<strap, eval=FALSE>>=
install.packages(strap)
library(strap)
geoscalePhylo(BeckLee_tree, cex.tip=0.8, cex.ts=0.6)
@

@@

The \dispRity package provides a \texttt{time.series} function that allows to split the morphospace time slices (using \texttt{method="continuous"}) or into time bins (using \texttt{method="discrete"}).
In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
Note that we can also provide here the table containing the first and last occurrences data for some fossils to take into account some fossils that might span between our different bins.

@@

The warning message tells us that some taxa were considered as single points in time because we didn't provide any temporal range information for some taxa.
The output object is a \dispRity object.
For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.
In brief however, \dispRity objects are lists of different elements (i.e. disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for visualisation reasons.

@@

Note that these objects will gradually contain more information when reaching higher steps in the disparity-through-time analysis.

\subsection{Bootstrapping the data}
Once we obtain our different subsamples, we might want to bootstrap or/and rarefy it (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems.
For that, we can use the \texttt{boot.matrix} function that allows to bootstrap the \dispRity object along with the \texttt{rarefaction} option that will rarefy it.

@@

## ... and display some more information on the previous step when called
boot_bin_morphospace
rare_bin_morphospace
@

@@

We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstrap/rarefaction).
Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.

In this example, we are going to calculate the spread of the data in each sub-sample by calculating the disparity as the sum of the variance of each dimension of the morphospace in each sub-sample using the \dispRity function.

@@

## Plotting the bootstrapped and rarefied results
plot(boot_disparity, main = "bootstrapped results")
plot(rare_disparity, main = "rarefied results")

@@

\subsection{Testing differences}
Finally, to draw some conclusions from these results, we need to apply some statistical testing.
As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
To do so, we can simply apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
Because our data is temporally auto-correlated (i.e. what happens in bin Y is dependent on what happened in bin X) and was pseudo-replicated (i.e. each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis.

>>>>

This demo showcases a typical disparity-through-time analyse: we are going to test whether the disparity changed through time in a subset of Eutheria since the last 100 million years.

@@

<<strap, eval=FALSE>>=
if(!require(strap)) install.packages("strap")
geoscalePhylo(BeckLee_tree, cex.tip=0.8, cex.ts=0.6)
@

@@

The \dispRity package provides a \texttt{time.series} function that allows to split the morphospace time slices (using \texttt{method = "continuous"}) or into time bins (using \texttt{method = "discrete"}).
In this example, we are going to split the morphospace into 5 equal time bins of 20 million years long from 100 million years ago (Mya) to the present.
Note that we will also provide to the function the table containing the first and last occurrences data for some fossils to take into account the fact that some fossils might span between our different time bins.

@@

The warning message tells us that some taxa were considered as single points in time because we didn't provide temporal range information for all the taxa: they are therefore considered as unique time points without any occurrence range.

The output object is a \dispRity object.
For details about this object class, please refer to the \href{https://github.com/TGuillerme/dispRity}{the \dispRity manual}.
In brief, \dispRity objects are lists of different elements (i.e. disparity results, morphospace sub-samples, etc...) that display only a summary of the object when calling the object for avoiding jamming the \R console.

@@

Note that these objects will gradually contain more information when reaching the following steps in the disparity-through-time analysis.

\subsection{Bootstrapping the data}
Once we obtain our different subsamples, we can bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrapping will allow to make each subsample more robust to outliers and the rarefaction will allow us to compare slices with the same number of taxa to get rid of eventual sampling problems (i.e. more taxa in one bin than the other).
The \texttt{boot.matrix} function that allows to bootstrap the \dispRity object and the \texttt{rarefaction} option within allows to rarefy it.

@@

## ... and display some more information on the previous step when printed:
boot_bin_morphospace
rare_bin_morphospace
@

@@

We can now calculate the disparity within each sub-sample along side with some confidence intervals generated by the pseudo-replication step above (bootstraps/rarefaction).
Disparity can be calculated in many ways and this package allows user to come up with their own disparity metrics.
For more info, please refer to the \href{https://github.com/TGuillerme/dispRity}{\dispRity manual and the metric vignette}.

In this example, we are going to calculate the spread of the data in each sub-sample by calculating the disparity as the sum of the variance of each dimension of the morphospace in each sub-sample using the \dispRity function.
Thus, in this example, the disparity is defined by the multi-dimensional variance of each sub-sample (i.e. the spread of the taxa within the morphospace).
Note that this metric comes with a caveat not solved here since it ignores the covariance among the dimensions of the morphospace.

@@

## Plotting the bootstrapped and rarefied results
plot(boot_disparity, type = "continuous", main = "bootstrapped results")
plot(rare_disparity, type = "continuous", main = "rarefied results")

@@

\subsection{Testing differences}
Finally, to draw some conclusions from these results, we need to apply some statistical testing.
As stated before, we wanted to test whether the disparity changed through time in this dataset since the last 100 million years.
To do so, we can apply some comparisons of the means between each time-bins in a sequential manner to see whether the disparity in bin X is equal to the disparity in bin Y which is in turn equal to the disparity in bin Z, etc.
Because our data is temporally auto-correlated (i.e. what happens in bin Y depends on what happened in bin X) and was pseudo-replicated (i.e. each bootstrap draw creates non-independent sub-samples because they are all based on the same sub-sample), we are going to apply a non-parametric mean comparison: the \texttt{wilcox.test}.
Also, we are going to apply a p-value correction to avoid type I error inflation due to performing multiple test for the same hypothesis (see \texttt{?p.adjust} for more details).
